Computing the bicharacter of huge subspaces of polynomials

### 2019-06-03 article/article.tex

\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{prop}{Proposition}
\newtheorem{lemme}{Lemma}
\newtheorem{coroll}{Corollary}
\newtheorem{definition}{Definition}

@@

	Therorem \ref{DiagHarm} has been proven for two sets of variables. It is not proven yet for more than two sets of variables but it has been verified by calculation for a large number of sets of variables so we assume it to be true since we will never reach that number of sets in our case.  \\

@@

	The polarization operators, ie the action of $\GLr$, commute with the action of $\Sn$. Therefore, we can compute independently each $\Sn$ isotypic component of $\harmonics_{n,r}$ by starting from a basis of harmonic polynomials that itself is adapted to its decomposition into isotypic components:
	$$\harmonics_{n}=:\bigoplus_\lambda \harmonics^\lambda\,.$$

	In practice, first we compute $\harmonics_{n}$ and decompose it into isotypic components using the Young symmetrizer. Then we use the polarization operators to obtain the isotypic components of $\harmonics_{n,r}$ from those of $\harmonics_{n}$. \\

	To define the Young symmetrizer, we first need to define two elements of the symmetric group algebra $\CC \Sn$. Let $P_{\lambda}$ be the subgroup of $\Sn$ of all row preserving permutations and $Q_{\lambda}$ the subgroup of the columns preserving permutations. Then we have $a_{\lambda} = \sum_{\sigma \in P_{\lambda}} \sigma$ and  $b_{\lambda} = \sum_{\sigma \in P_{\lambda}} \sgn(\sigma) \sigma$. Finally, the Young symmetrizer is the element $$c_{\lambda} = a_{\lambda} b_{\lambda}.$$

@@

	This theorem guarantees that $c_{\lambda}$ is an isotypic projector and that applying it on the elements of $\harmonics_{n}$ gives representatives of the isotypic components of $\Sn$. \pauline{Do we need to prove explicitly commutation with symmetric operators?}

	As mentioned above, since the action of $\GLr$ and $\Sn$ commutes, at this stage we only have to apply the polarization operators on each isotypic component independently to be able to compute the bicharacter. \\

@@

	To conclude, it is sufficient to compute $\harmonics_k^P$ for one tableau of each shape $\lambda$. This reduces the complexity by the square of the number of standard tableaux of shape $\lambda$

@@

	But that's mostly because they are antisymmetric w.r.t. a large column stabilizer $S_P$. It's therefore sufficient to keep one representative monomial per $S_P$ orbit (and do some appropriate straightening whenever a new monomial is created).

@@

	**Gain**: in theory up to $n!$. In practice between one and two orders of magnitude due to the overhead of straightening monomials.

@@

	**Gain**: an order of magnitude; below are the total dimension of each $\harmonics_5^\lambda$ and its restriction to homogeneous components of decreasing multidegree:

@@

	**Current gain:** another order of magnitude; down from 3 days to 45 minutes for $\harmonics_5^{1^6}$

	**Potential gain:** another order of magnitude, both in time and memory:

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{prop}{Proposition}
\newtheorem{lemme}{Lemma}
\newtheorem{coroll}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

	Therorem \ref{DiagHarm} has been proven for two sets of variables. It is not proven yet for more than two sets of variables but it has been verified by calculation by Mike Zabrocki for a large number of sets of variables so we assume it to be true (since we will never reach that number of sets in our case).  \\

@@

	First of all, we need an efficient way to compute the representations of $\Sn$. To do that, we will use an isotypic projector to obtain the isotypic components of $\Sn$ and from that we will be able to recover the irreducible characters of $\Sn$ and their multiplicities. 
	The isotypic projector we use is called the Young symmetrizer and is define as follow.

	We first need to define two elements of the symmetric group algebra $\CC \Sn$. Let $P_{\lambda}$ be the subgroup of $\Sn$ of all row preserving permutations of $\lambda$ and $Q_{\lambda}$ the subgroup of the columns preserving permutations of $\lambda$. Then we have $a_{\lambda} = \sum_{\sigma \in P_{\lambda}} \sigma$ and  $b_{\lambda} = \sum_{\sigma \in P_{\lambda}} \sgn(\sigma) \sigma$. Finally, the Young symmetrizer is the element $$c_{\lambda} = a_{\lambda} b_{\lambda}.$$

@@

	This theorem guarantees that $c_{\lambda}$ is an isotypic projector and that applying it on the elements of $\harmonics_{n}$ gives representatives of the isotypic components of $\Sn$. \pauline{Do we need to prove explicitly commutation with symmetric operators?}  \\

	The action of $\GLr$ commutes with the action of $\Sn$. Therefore, we can compute independently each $\Sn$ isotypic component of $\harmonics_{n,r}$ by starting from a basis of harmonic polynomials that itself is adapted to its decomposition into isotypic components:
	$$\harmonics_{n}=:\bigoplus_\lambda \harmonics^\lambda\,.$$

	In practice, since the polarization operators commute with the Young symmetrizer, we first compute $\harmonics_{n}$ and decompose it into isotypic components using the Young symmetrizer. Then we use the polarization operators to obtain the isotypic components of $\harmonics_{n,r}$ from those of $\harmonics_{n}$. \\

@@

	To conclude, it is sufficient to compute $\harmonics_k^P$ for one tableau of each shape $\lambda$. This reduces the complexity by the square of the number of standard tableaux of shape $\lambda$.

@@

	But that's mostly because they are anti-symmetric w.r.t. a large column stabilizer $S_P$. It's therefore sufficient to keep one representative monomial per $S_P$ orbit (and do some appropriate straightening whenever a new monomial is created).

	Inside isotypic component, polynomials are anti-symmetric with respect to the type $\lambda$ of the component. This is due to the fact that applying the young symmetrizer $y_{\lambda}$ firt symmetrize the polynomial with respect to the row preserving permutations of $\lambda$ and then anti-symmetrize the result with respect to the column preserving permutations of $\lambda$.

	Thus it is sufficient to keep only a reduced form of the polynomials.

	\begin{example}
		$\Delta_{21} = -x_{2} \theta_{1} + x_{3} \theta_{1} + x_{1} \theta_{2} - x_{3} \theta_{2} - x_{1} \theta_{3} + x_{2} \theta_{3}$ is type $111$.
		Reduce form : $x_1\theta_2$. \\
		$\partial_{x_2}\Delta_{21} = -\theta_1 + \theta_3$ is type $21$. 
		Reduce form : $-\theta_1$ 
	\end{example}

@@

	\textbf{Gain:} in theory up to $n!$. In practice between one and two orders of magnitude due to the overhead of straightening monomials.

@@

	\begin{example}
		$x_1^2 - 2x_1x_2 + 2x_2x_3 - x_3^2 \in \mathcal{H}_{3}$ so $y_1^2 - 2y_1y_2 + 2y_2y_3 - y_3^2 \in \mathcal{H}_{3}$

		$-2x_{1} y_{1} y_{2} + x_{1} y_{2}^{2} \in \mathcal{H}_{3}$ so $-2y_{1} x_{1} x_{2} + y_{1} x_{2}^{2} \in \mathcal{H}_{3}$
	\end{example}

	\textbf{Gain:} an order of magnitude; below are the total dimension of each $\harmonics_5^\lambda$ and its restriction to homogeneous components of decreasing multidegree.

@@

	\textbf{Current gain:} another order of magnitude; down from 3 days to 45 minutes for $\harmonics_5^{1^6}$

	\textbf{Potential gain:} another order of magnitude, both in time and memory.
